---
title: "Activity9"
format: html
warning: false
message: false
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```



## Model Selection & Diagnostic Criteria

Today, we will walk through a complete model selection exercise using VIX data. Our goals are to:

- Perform Exploratory Data Analysis (EDA) with visual diagnostics.
- Fit built-in models (ARIMA and ETS) using the `Tidyverts` framework.
- Evaluate models based on five metrics: AIC, AICc, BIC, Mean Squared Error (MSE), and Root Mean Squared Error (RMSE).
- Conclude on the optimal model by comparing these metrics along with residual diagnostics and forecast performance.

For a model with $n$ observations and $k$ parameters (including the intercept), the evaluation criteria are defined as:

$$
\begin{align}
\text{AIC} &= n\ln\left(\frac{\text{RSS}}{n}\right) + 2k,\\[1mm]
\text{BIC} &= n\ln\left(\frac{\text{RSS}}{n}\right) + k\ln(n),\\[1mm]
\text{AICc} &= \text{AIC} + \frac{2k(k+1)}{n-k-1},
\end{align}
$$

and the error metrics:

$$
\begin{align}
\text{MSE} &= \frac{\text{RSS}}{n},\\[1mm]
\text{RMSE} &= \sqrt{\text{MSE}}.
\end{align}
$$


# Data Preparation

We retrieve the VIX data and format as a `tsibble`, handling missing values while preserving the original adjusted price series.

```{r}
# Retrieve VIX data from 2000-01-01 to current date and convert to a tsibble
vix_data <- tq_get("^VIX", from = "2000-01-01", 
                   to = Sys.Date(), 
                   get = "stock.prices") %>% 
  as_tsibble(index = date)

# Clean and interpolate missing values
vix_data <- vix_data %>% 
  tsibble::fill_gaps() %>% 
  mutate(adjusted = na.approx(adjusted)) %>% 
  select(date, adjusted) %>% 
  tidyr::drop_na()
```

# Exploratory Data Analysis (EDA)

## Time Series Plot

```{r}
vix_data %>% 
  autoplot(adjusted) +
  labs(title = "VIX Adjusted Prices", y = "Adjusted", x = "Date")
```

## ACF and PACF Analysis

```{r}
vix_data %>% 
  ACF(adjusted) %>% 
  autoplot() +
  labs(title = "ACF of VIX Data")

vix_data %>% 
  PACF(adjusted) %>% 
  autoplot() +
  labs(title = "PACF of VIX Data")
```

> **Observation:**  
> The persistence in ACF suggests potential need for differencing, but we'll let ARIMA automatically handle any required transformations through its $d$ parameter.

# Model Fitting Using Tidyverts

## 1. ARIMA Model (Original Series)

```{r}
# Fit an ARIMA model on original series
model_arima <- vix_data %>% 
  model(ARIMA(adjusted))

# Display summary report
report(model_arima)

# Display accuracy metrics
model_arima %>% accuracy()
```

## 2. ETS Model

```{r}
# Fit ETS model on original series
model_ets <- vix_data %>% 
  model(ETS(adjusted))

# Display summary report
report(model_ets)

# Display accuracy metrics
model_ets %>% accuracy()
```

# Forecasting with the Optimal Model

```{r}
# Forecast 12 months ahead using ARIMA
forecast_arima <- model_arima %>% 
  forecast(h = "12 months")

# Plot forecast in original units
forecast_arima %>% 
  autoplot(vix_data) +
  labs(title = "12-Month VIX Forecast", y = "Adjusted VIX", x = "Date")
```


## **Lab Activity: Modeling Stock Returns with ARIMA**

Model daily SPY returns using ARIMA, analyze residuals, and compare with ETS. 

**Tasks**  

1. Retrieve SPY data (2015-01-01 to present) and compute **daily log returns**  
2. Perform EDA (time series plot, ACF/PACF)  
3. Fit ARIMA(2,0,1) and ETS models to returns  
4. Compare models using AIC and RMSE  
5. Forecast 1-month ahead returns

<!--

**Solution**

```{r}
# Task 1: Data Preparation
spy_data <- tq_get("SPY", from = "2015-01-01") %>% 
  as_tsibble(index = date) %>% 
  tsibble::fill_gaps() %>% 
  mutate(adjusted = na.approx(adjusted),
         log_return = difference(log(adjusted))) %>% 
  select(date, adjusted, log_return) %>% 
  tidyr::drop_na()


# Task 2: EDA
spy_data %>% autoplot(log_return) +
  labs(title = "SPY Daily Log Returns", y = "Return")

spy_data %>% ACF(log_return) %>% autoplot()
spy_data %>% PACF(log_return) %>% autoplot()

# Task 3: Model Fitting
fit_arima <- spy_data %>% 
  model(ARIMA(log_return ~ pdq(2,0,1)))

fit_ets <- spy_data %>% 
  model(ETS(log_return))

# Task 4: Model Comparison
glance(fit_arima) %>% select(AIC, AICc, BIC)
glance(fit_ets) %>% select(AIC, AICc, BIC)

accuracy(fit_arima)
accuracy(fit_ets)

# Task 5: Forecasting
fit_arima %>% 
  forecast(h = 30) %>% 
  autoplot(spy_data) +
  labs(title = "30-Day SPY Return Forecast", y = "Log Return")
```

-->

