---
title: "Activity9"
format: html
warning: false
message: false
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
library(vars)
library(COVID19)
library(GGally)
select <- dplyr::select
```


## COVID Data Analysis and Granger causality test

### Objective

Investigate whether changes in COVID-19 vaccination numbers help predict changes in confirmed cases using a Granger causality test.


```{r}
# Retrieve COVID-19 data for the United States and prepare a tsibble
covid_data <- covid19(verbose = FALSE) %>% 
  filter(administrative_area_level_1 == "United States") %>% 
  mutate(date = ymd(date)) %>% 
  as_tsibble(index = date) %>% 
  select(date, confirmed, vaccines) %>% 
  mutate(vaccines = as.integer(vaccines)) %>% 
  drop_na()
```


```{r}
# Compute daily changes in confirmed cases and vaccinations
covid_data <- covid_data %>% 
  mutate(dConfirmed = difference(confirmed),
         dVacc = difference(vaccines)) %>% 
  drop_na()
```


```{r}
# Plot daily changes in confirmed cases and vaccinations
covid_data %>% 
  autoplot(vars(dConfirmed, dVacc)) +
  labs(title = "Daily Changes: Confirmed Cases & Vaccinations", x = "Date")
```


### Granger Causality Test: Theory & Implementation

The Granger causality test checks whether past values of one variable (here, dVacc) provide statistically significant information for predicting another variable (dConfirmed).

**Implications and Uses:**

- A significant result (p-value < 0.05) suggests that changes in vaccinations Granger-cause changes in confirmed cases.
- This does not imply true causality, but indicates that past vaccination data improves forecast accuracy for confirmed cases.
- Such insights can help inform public health policies by highlighting the predictive value of vaccination trends.

**P-value Interpretation:**

- p < 0.05: Reject the null hypothesis. Past vaccination changes significantly improve prediction of confirmed cases.
- p ≥ 0.05: Fail to reject the null hypothesis. Vaccination changes do not add significant predictive power.


### VAR Modeling and IRF Computation

```{r}
# Test if changes in vaccinations (dVacc) Granger-cause changes in confirmed cases (dConfirmed) using a lag of 1.
model_full <- lm(dConfirmed ~ lag(dVacc, 1), data = covid_data)
model_restricted <- lm(dConfirmed ~ 1, data = covid_data)
F_stat <- ((deviance(model_restricted) - deviance(model_full)) / 1) / 
          (deviance(model_full) / (nrow(covid_data) - 2 - 1))
df1 <- 1
df2 <- nrow(covid_data) - 2 - 1
p_value <- pf(F_stat, df1, df2, lower.tail = FALSE)
p_value
```


## Lab Activity: Exploring Granger Causality with Different Lags

**Task:** Using the covid_data, perform Granger causality tests with lag orders 1, 2, 3, 5, and 10.

For each lag, compute the p-value to assess whether lagged vaccination changes improve the prediction
of confirmed cases. Answer the following:

1. How do the p-values change with different lag orders?
2. Which lag order appears most appropriate for capturing the predictive relationship?

Write a brief discussion on how these findings might inform public health policy.

<!--


```{r}
# Test Granger causality using multiple lag orders and collect the results.
lag_orders <- c(1, 2, 3, 5, 10)
results <- data.frame(Lag = integer(), P_Value = numeric())

for (lag_order in lag_orders) {
  model_full <- lm(dConfirmed ~ lag(dVacc, lag_order), data = covid_data)
  model_restricted <- lm(dConfirmed ~ 1, data = covid_data)
  F_stat <- ((deviance(model_restricted) - deviance(model_full)) / 1) / 
            (deviance(model_full) / (nrow(covid_data) - lag_order - 1))
  df1 <- 1
  df2 <- nrow(covid_data) - lag_order - 1
  p_value <- pf(F_stat, df1, df2, lower.tail = FALSE)
  results <- rbind(results, data.frame(Lag = lag_order, P_Value = p_value))
}

print(results)
```

-->
