---
title: "Activity23"
format: html
warning: false
message: false
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```


## **1. SARIMA Model Structure**  

A SARIMA$(p,d,q)(P,D,Q)_m$ model combines:  

- **Regular components**: AR(p), MA(q) terms for short-term patterns  
- **Seasonal components**: Seasonal AR(P), MA(Q) terms at period $m$  
- **Differencing**: $d$ regular differences + $D$ seasonal differences  

## **2. Strategic Differencing**  

**Key Principle**: Use minimal differencing to stabilize mean/variance  

```{r}
# AirPassengers dataset
ap_ts <- tsibble::as_tsibble(AirPassengers) %>% 
  index_by(Date = yearmonth(index)) %>% 
  rename(Passengers = value)

# Automated selection
ap_ts %>% 
  features(Passengers, list(unitroot_kpss, unitroot_ndiffs, unitroot_nsdiffs))

# Visual check
ap_ts %>%
  gg_tsdisplay(difference(log(Passengers), lag = 12))
```

**Insight**: Seasonal differencing (lag=12) removes yearly patterns while preserving monthly trends  


## **3. Model Building**  

### **3.1 Candidate Models**  

```{r}
models <- ap_ts %>%
  model(
    Auto = ARIMA(log(Passengers)),
    Manual1 = ARIMA(log(Passengers) ~ pdq(1,1,1) + PDQ(0,1,1, period=12)),
    Manual2 = ARIMA(log(Passengers) ~ pdq(2,1,0) + PDQ(1,1,0, period=12))
  )

glance(models) %>% arrange(AICc)
```

### **3.2 Coefficient Check**  

For top model:  

```{r}
final_model <- models %>% 
  select(Auto)

tidy(models) %>% 
  filter(.model == "Auto") 
```


## **4. Model Refinement Cycle**  

1. Start with automatic differencing  
2. Compare multiple model specifications  
3. Validate residuals systematically  
4. Iterate using PACF patterns  

```{r}
# Final refinement example
ap_ts %>%
  model(
    Best = ARIMA(log(Passengers) ~ pdq(1,1,1) + PDQ(0,1,1, period=12))
  ) %>% 
  report()
```


## **Lab Activity**

## **1. Data Preparation**

```{r}
library(fpp3)
accidents <- as_tsibble(USAccDeaths) %>% 
  rename(Deaths = value) %>% 
  mutate(Month = yearmonth(index))
```

**Q1:** Examine the seasonal patterns using `gg_season()`. What type of seasonality dominates this series?  


## **2. Model Specification**

Fit SARIMA(1,1,1)(0,1,1)₁₂ model with maximum likelihood estimation

```{r}
sarima_fit <- accidents %>%
  model(SARIMA = ARIMA(Deaths ~ pdq(1,1,1) + PDQ(0,1,1, period = 12),
                      stepwise = FALSE, approximation = FALSE))
```

**Q2:** Interpret the model structure:  

a) What does the (1,1,1) non-seasonal component represent?  

b) Why do we use PDQ(0,1,1) for seasonal terms?  



## **3. Parameter Estimation**

**Task:** Extract and interpret coefficients

```{r}
tidy(sarima_fit) %>% 
  mutate(signif = ifelse(p.value < 0.05, "***", ""))
```

**Q3:** Which coefficients are statistically significant (α=0.05)? What does the MA(1) coefficient suggest?  

## **4. Residual Diagnostics**

### **4.1 Visual Analysis**

```{r}
sarima_fit %>% 
  augment() %>% 
  ACF(.innov) %>% 
  autoplot() +
  geom_hline(yintercept = c(-1,1)*1.96/sqrt(nrow(accidents)), linetype=2)
```

**Q4:** Do residuals show concerning autocorrelation patterns? Justify your answer.  

### **4.2 Formal Tests**

```{r}
augment(sarima_fit) %>% 
  features(.innov, list(
    LB_test = ~ljung_box(.x, lag=24, dof=4)
  ))
```

**Q5:** Interpret the Ljung-Box (LB) test results: 

a) Can we maintain the white noise assumption? 


## **5. Model Validation**

**Task:** Check specification robustness

```{r}
accidents %>% 
  model(
    SARIMA_CSS = ARIMA(Deaths ~ pdq(1,1,1) + PDQ(0,1,1), 
                      method = "CSS-ML"),
    AutoSARIMA = ARIMA(Deaths)
  ) %>% 
  glance()
```

**Q6:** Compare AICc values across estimation methods. Does our original model remain preferred?  
