---
title: "Activity20"
format: html
warning: false
message: false
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```


**Cross-Correlation & Multiserial Dynamics**  

*Objective: Analyze lead-lag relationships between series*

The cross-correlation function at lag k is given by:

$$
\rho_{XY}(k) = \frac{\gamma_{XY}(k)}{\sqrt{\gamma_{XX}(0)\,\gamma_{YY}(0)}}
$$

where:

 - $\gamma_{XY}(k)$ is the cross-covariance at lag k between series X and Y
 - $\gamma_{XX}(0)$ and $\gamma_{YY}(0)$ are the variances (covariance at lag 0)
   of X and Y, respectively
 - $\rho_{XY}(k)$ thus ranges between -1 and +1



### How to interpret the CCF?

The CCF at lag k measures how strongly the current values of one series correlate with the values of the other series shifted by k time steps.

- Positive lag (k > 0) indicates how much the first series LEADS the second.
- Negative lag (k < 0) indicates how much the second series LEADS the first.
- A high absolute value of CCF at a particular lag suggests a strong
 lead-lag relationship at that time shift.


### 1.1 CCF Analysis with `pelt`

```{r}
pelt |> 
  as_tsibble() |>
  CCF(Lynx, Hare) |> 
  autoplot() +
  labs(title = "Hare peaks lead Lynx by 1-2 years (predator-prey dynamics)")
```


- A single cross-correlation plot is sufficient for understanding how two time series move relative to each other over different time shifts.

- Peaks at positive lags suggest that the first series' changes come before changes in the second series (leading). Peaks at negative lags suggest the second series leads the first.

- In ecological contexts (like hare vs. lynx), a positive lag at which the CCF is high might mean prey population spikes occur 1-2 years before predator population spikes.


### 1.2 Spurious Correlation Caveat

```{r}
# Simulate independent series
set.seed(123)
fake_data <- tsibble(t=1:100, x=rnorm(100), y=rnorm(100), index=t)
fake_data |> 
  CCF(x,y) |> autoplot()
```

**Critical Thinking**: Even independent series may show "significant" correlations by chance. Always validate with domain knowledge.


**Simulation & Model Identification**  

*Hands-on ARMA process experimentation*

### 1.3 AR(2) Simulation & Diagnostics

```{r}
sim_ar2 <- arima.sim(n=200, list(ar=c(0.6, -0.3))) 

sim_ar2 |> 
  as_tsibble() |>
  ACF() |> 
  autoplot() # Decaying oscillations

sim_ar2 |> 
  as_tsibble() |>
  PACF() |> 
  autoplot() # Spikes at lags 1-2
```

### 1.4 MA(1) Characteristics

```{r}
sim_ma1 <- arima.sim(n=200, list(ma=0.8))

sim_ma1 |> 
  as_tsibble() |>
  ACF() |> 
  autoplot() # Cutoff after lag 1

sim_ma1 |> 
  as_tsibble() |>
  PACF() |> 
  autoplot() # Exponential decay
```

**Golden Rule**:  

- AR(p): PACF significant for first p lags  
- MA(q): ACF significant for first q lags  
- ARMA: Both decay gradually
