---
title: "Activity37"
format: 
  live-html:
    theme:
      light: [lux, theme-light.scss]
      dark: [superhero, theme-dark.scss] 
engine: knitr
toc: true
---

{{< include _extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
library(vars)
library(tseries)
library(tsDyn)
library(quantmod)
select <- dplyr::select
```


#### **Modeling Inflation Dynamics**  

Let's model inflation ($\pi_t$) with Unemployment, GDP growth, and CPI (lagged). The relationship can be written as:  
$$\pi_t = \beta_0 + \beta_1 \text{Unemployment}_t + \beta_2 \text{CPI}_{t-1} + \beta_3 \text{GDP\_growth}_t + \epsilon_t$$  
There are a lot of dynamics at play here. GDP growth ($\text{GDP\_growth}$) may signal demand-pull inflationary pressures, while unemployment ($\text{Unemployment}_t$) and lagged CPI ($\text{CPI}_{t-1}$) capture labor market slack and inertial price effects. However, structural breaks—such as shifts in productivity regimes, fiscal/monetary policy pivots, or global supply chain disruptions—can alter the **sign**, **magnitude**, or **significance** of these relationships over time.  

1. **Detect breaks in all coefficients**: Do a Chow test to identify shifts in $\beta_1$ (unemployment’s impact), $\beta_2$ (CPI persistence), *and* $\beta_3$ (GDP growth sensitivity). For example, a break in $\beta_3$ might reflect a transition from a high-growth, high-inflation era to a stagflationary regime.  
2. **Segmented regression with regime-specific GDP effects**: Estimate $$Y_t = \beta_0^{(k)} + \beta_1^{(k)}\text{Unemployment}_t + \beta_2^{(k)}\text{CPI}_{t-1} + \beta_3^{(k)}\text{GDP\_growth}_t + \epsilon_t^{(k)}$$ for each regime $k$. This reveals whether GDP growth amplifies inflation in certain eras (e.g., post-recession recoveries) but not others (e.g., during secular stagnation).  
3. **STL decomposition**: Separate the trend component of inflation to distinguish whether GDP-driven fluctuations are cyclical (e.g., business cycle upswings) or structural (e.g., sustained growth-inflation linkages post-break).  



```{r}
getSymbols("CPIAUCSL", src = "FRED", auto.assign = TRUE)

cpi_data <- tibble(date = index(CPIAUCSL), CPI = as.numeric(CPIAUCSL[,1])) %>%
  arrange(date) %>%
  mutate(inflation = 100 * (CPI / lag(CPI) - 1)) %>%
  mutate(Quarter = yearquarter(date)) %>% 
  as_tsibble(index = date) %>% 
  index_by(Quarter) %>% 
    summarize(CPI = mean(CPI),
              inflation = mean(inflation))

# Get US unemployment (monthly)
getSymbols("UNRATE", src = "FRED", auto.assign = TRUE)
us_unemp <- data.frame(date = index(UNRATE), 
                      Unemployment = as.numeric(UNRATE$UNRATE)) %>%
  as_tsibble(index = date) %>% 
  mutate(Quarter = yearquarter(date)) %>% 
  index_by(Quarter) %>% 
    summarize(Unemployment = mean(Unemployment))


# Get US GDP (quarterly) and merge
getSymbols("GDP", src = "FRED", auto.assign = TRUE)
us_gdp <- data.frame(date = index(GDP), GDP = as.numeric(GDP$GDP)) %>%
  as_tsibble(index = date) %>% 
  mutate(Quarter = yearquarter(date))

combined_data <- us_gdp %>% 
  inner_join(us_unemp, by = "Quarter") %>% 
  inner_join(cpi_data, by = "Quarter") %>% 
  mutate(GDP_growth = 100 * (GDP / lag(GDP) - 1)) %>% 
  dplyr::select(Quarter, date, GDP, GDP_growth, Unemployment, CPI, inflation) %>% 
  drop_na()

# View the combined data
head(combined_data) %>% knitr::kable()
```


```{r}
# Chow test for structural breaks
break_test <- Fstats(inflation ~ GDP_growth + Unemployment + CPI, data = combined_data)
plot(break_test)  # Identify potential breakpoints

# Estimate breakpoint date(s)
bp <- breakpoints(inflation ~  GDP_growth + Unemployment + CPI, data = combined_data)

combined_data$Quarter[bp$breakpoints] 

# Fit segmented regression
segmented_model <- lm(inflation ~ GDP_growth + Unemployment + CPI, 
                      data = combined_data, 
                      subset = breakpoints(bp)$breakpoints)
```


```{r}
# Get breakpoint indices 
bp_indices <- bp$breakpoints 

# Split data into segments
segments <- split(combined_data, 
                 findInterval(1:nrow(combined_data), 
                              vec = c(0, bp_indices, nrow(combined_data))))

# Fit models to each segment
segment_models <- lapply(segments, function(df) {
  lm(inflation ~ GDP_growth + Unemployment + CPI, data = df)
})

# Combine residuals from all segments
all_residuals <- unlist(lapply(segment_models, residuals))
```


```{r}
# Plot segmented regressions
ggplot(combined_data, aes(y =inflation, x=Quarter)) +
  geom_point() +
  geom_smooth(data = segments[[1]], method = "lm", se = FALSE, color = "red") +
  geom_smooth(data = segments[[2]], method = "lm", se = FALSE, color = "blue") +
  labs(title = "Segmented Regression by Breakpoints")
```


```{r}
# Inflation analysis
combined_data %>%
  as_tsibble(index = Quarter) %>%
  model(STL(inflation ~ trend() + season())) %>%
  components() %>%
  autoplot() + labs(title = "Inflation Decomposition (STL)")
```


```{r}
lapply(segment_models, summary)  # Show coefficients for each regime
```



### **Activities** 

#### **Parameter Stability Assessment**

**Q1: Test if the employment-inflation relationship remains stable using recursive residuals**



```{r}
library(dynlm)
rolling_residuals <- efp(inflation ~ GDP_growth  + Unemployment + CPI, data = combined_data, type = "Rec-CUSUM")

# Plot and interpret:
plot(rolling_residuals)
abline(h = 1.358, lty = 2)  # 95% significance boundary
```

**Task**: Explain what crossing the dashed line implies about parameter stability. How does this complement the Chow test results?

---

#### **Time-Varying Volatility Analysis**

**Q2: Model volatility changes using GARCH(1,1)**


```{r}
library(rugarch)
garch_spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                         mean.model = list(armaOrder = c(1,1)))

garch_fit <- ugarchfit(garch_spec, data = combined_data$inflation)

# Plot conditional volatility
plot(garch_fit, which = 3)
```

**Task**: Identify periods of high volatility clustering. Could these coincide with structural breaks in mean relationships?



#### **Rolling Window Analysis**

**# Q3: Calculate 5-year rolling correlations**

```{r}
library(slider)
combined_data <- combined_data %>%
  mutate(
    rolling_cor = slide_dbl(
      .x = cur_data(),
      .f = ~ cor(.x$inflation, .x$Unemployment),
      .before = 60  # 5-year window
    )
  )

# Visualize
ggplot(combined_data, aes(Quarter, rolling_cor)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed")
```

**Task**: When does employment become negatively correlated with inflation? How does this relate to Phillips curve theory? The **Phillips Curve** is an economic theory describing an inverse relationship between unemployment and inflation, first observed by A.W. Phillips (1958) using UK wage data.


