---
title: "Activity10"
format: html
warning: false
message: false
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```


# Multiple Regression & Collinearity

Multiple regression models the relationship between a dependent variable and multiple independent variables. In many real‐world scenarios, predictors can be correlated—this is known as **multicollinearity**. High multicollinearity inflates the variance of the coefficient estimates, making them unstable and difficult to interpret. A common diagnostic measure is the **Variance Inflation Factor (VIF)**. For a regression model:

$$
\begin{align}
Y_t &= \beta_0 + \beta_1 X_{1,t} + \beta_2 X_{2,t} + \varepsilon_t,
\end{align}
$$

the VIF for predictor $\\(X_i\\)$ is computed as:

$$
\text{VIF}(X_i) = \frac{1}{1 - R_i^2},
$$

where $\\(R_i^2\\)$ is the coefficient of determination when $\\(X_i\\)$ is regressed on the other predictors. VIF values greater than 5 (or 10, depending on context) indicate potentially problematic collinearity.


## Real-World Examples

### 1. Economy Example (US Macro Data)

*Dataset:* **us_change** (available in the `fpp3` package)  
*Model:*  

$$
\text{Consumption}_t = \beta_0 + \beta_1 \text{Income}_t + \beta_2 \text{Production}_t + \varepsilon_t
$$

In this example, we explore how Income and Production drive Consumption. We then use VIF to diagnose multicollinearity.

```{r economy_example, message=FALSE, warning=FALSE}
library(car)

# Fit the TSLM model on us_change data
model_macro <- us_change %>%
  model(tslm_macro = TSLM(Consumption ~ Income + Production))

# Display model coefficients
model_macro %>%
  tidy(tslm_macro) %>%
  knitr::kable(caption = 'US Macro Model Coefficients')

# Compute VIF values using a standard lm object
macro_lm <- lm(Consumption ~ Income + Production, data = us_change)
vif_macro <- car::vif(macro_lm)
vif_macro
```

### 2. COVID Example (Mobility)

*Dataset:* **oxcgrt** (Oxford COVID Policy Tracker)  

*Model:*  

$$
\log(\text{ConfirmedCases}_t) = \beta_0 + \beta_1 \text{ConfirmedDeaths}_t + \beta_2 \text{StringencyIndexAverage}_t + \varepsilon_t
$$

Here, we investigate the effect of COVID-19 deaths and policy stringency on the growth of confirmed cases, using a log-transformation to account for exponential growth. VIF is computed to check for multicollinearity between the predictors.

```{r covid_example, message=FALSE, warning=FALSE}
library(readr)
url <- "https://github.com/OxCGRT/covid-policy-dataset/raw/main/data/OxCGRT_compact_national_v1.csv"
oxcgrt <- read_csv(url)

# Filter for the United States and create a log-transformed 'ConfirmedCases' variable
oxcgrt_us <- oxcgrt %>% 
  filter(CountryName == 'United States') %>% 
  mutate(Date = ymd(Date),
         log_ConfirmedCases = log(1 + ConfirmedCases)) %>% 
  as_tsibble(index = Date) %>% 
  drop_na(log_ConfirmedCases, ConfirmedDeaths, StringencyIndex_Average)

# Fit the TSLM model
model_covid <- oxcgrt_us %>%
  model(tslm_covid = TSLM(log_ConfirmedCases ~ ConfirmedDeaths + StringencyIndex_Average))

# Display model coefficients
model_covid %>% 
  tidy(tslm_covid) %>%
  knitr::kable(caption = 'COVID Model Coefficients')

# Compute VIF values using a standard lm object (car::vif works with lm)
covid_lm <- lm(log_ConfirmedCases ~ ConfirmedDeaths + StringencyIndex_Average, data = as_tibble(oxcgrt_us))
vif_covid <- car::vif(covid_lm)
vif_covid
```


### Lab Activity 1: Assessing Multicollinearity in US Macro Data

**Prompt:**  

1. Using the **us_change** dataset, fit a TSLM model with Consumption as the response and all of the remaining variables as predictors.  
2. Compute the VIF for each predictor using the `car::vif()` function.  
3. Interpret the VIF values and discuss whether there is evidence of multicollinearity.

<!--

**Solution:**

```{r lab_activity1, message=FALSE, warning=FALSE}
# Fit the TSLM model on us_change data
model_macro_lab <- us_change %>%
  model(tslm_macro_lab = TSLM(Consumption ~ .))

# Convert to an lm object for VIF computation
macro_lm_lab <- lm(Consumption ~ ., data = us_change)
vif_macro_lab <- car::vif(macro_lm_lab)
vif_macro_lab
```

-->

### Lab Activity 2: Exploring COVID Policy Measures and Their Effects

**Prompt:**  

1. Using the **oxcgrt** dataset for the United States, create a new variable `log_ConfirmedDeaths` as $\\(\\log(1 + \text{ConfirmedDeaths})\\)$.  
2. Fit a TSLM model with `ConfirmedCases`, `StringencyIndex_Average`, `C6M_Stay at home requirements` as predictors for `log_ConfirmedDeaths`.  
3. Compute the VIF for the predictors and interpret the results in terms of potential collinearity.


<!--

**Solution:**

```{r lab_activity2, message=FALSE, warning=FALSE}
# Filter and transform the data
oxcgrt_lab <- oxcgrt %>% 
  filter(CountryName == 'United States') %>% 
  mutate(Date = ymd(Date),
         log_ConfirmedDeaths = log(1 + ConfirmedDeaths)) %>% 
  as_tsibble(index = Date) %>% 
  drop_na(ConfirmedCases, log_ConfirmedDeaths, StringencyIndex_Average)

# Fit the TSLM model
model_covid_lab <- oxcgrt_lab %>%
  model(tslm_covid_lab = TSLM(log_ConfirmedDeaths ~ ConfirmedCases + StringencyIndex_Average + `C6M_Stay at home requirements`))

# Convert to an lm object for VIF computation
covid_lm_lab <- lm(log_ConfirmedDeaths ~ ConfirmedCases + StringencyIndex_Average + `C6M_Stay at home requirements`, data = as_tibble(oxcgrt_lab))
vif_covid_lab <- car::vif(covid_lm_lab)
vif_covid_lab
```

-->

