---
title: "Activity1"
format: live-html
engine: knitr
toc: true
webr:
  packages:
    - dplyr
    - ggplot2
    - purrr
    - feasts
    - fpp3
    - fable
    - astsa
warning: false
message: false
---

{{< include _extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
# Time Series Essentials
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```


## Introduction to Applied Time Series

Time series analysis involves studying data points collected or recorded over time. Its applications range from economics to biology, where understanding patterns, trends, and predicting future events is crucial.

Mathematically, a time series $\left\{X_t\right\}$ is a sequence of random variables indexed by time $t$. A central concept is stationarity. A time series is (weakly) stationary if its statistical properties do not change over time.

Formally, $\left\{X_t\right\}$ is weakly stationary if:
$\mathbb{E}\left[X_t\right]=\mu, \quad \operatorname{Var}\left(X_t\right)=\sigma^2, \quad$ and $\quad \operatorname{Cov}\left(X_t, X_{t+k}\right)=\gamma_k$ for all $t$.
Stationarity is crucial for reliable modeling and forecasting, as many models assume constant mean and variance over time.

## Recap: Multiple Linear Regression

Multiple linear regression models a response variable $Y$ as a linear function of predictors $X_1, X_2, \ldots, X_p:$

$$
Y=\beta_0+\beta_1 X_1+\beta_2 X_2+\cdots+\beta_p X_p+\varepsilon
$$

where $\varepsilon$ is the error term.
Regression analysis serves as a foundation for understanding relationships between variables, which extends into time series when considering trends and seasonality.

## Using `Tidyverts` in the Tidyverse Ecosystem

The `Tidyverts` collection integrates time series analysis into the tidyverse framework, allowing seamless data manipulation and visualization with familiar syntax.

## Understanding Time Series Data

Time series data for a single entity over time can be represented as a `tsibble`, which is a tidy data structure that inherently understands time.

```{r}
example1 <- tsibble(
   year = 2015:2019,         # <1>
   y = c(123, 39, 78, 52, 110),  # <2>
   index = year              # <3>
)

str(example1)                 # <4>
```

1. Define a time index from 2015 to 2019.
2. Create corresponding observations for each year.
3. Construct a `tsibble` with year as the index.
4. Use `str()` to inspect the structure of the `tsibble.`

## Working with Real Datasets

### Example: Google Stock Data

```{r}
#| label: fig-intro-googleclose
#| fig-cap: "Daily closing stock prices of Google (in USD)"
google <- read_csv("data/google.csv")        # <1>
google %>%  
  as_tsibble() %>%                                   # <2>
  autoplot(Close) +                                 # <3>
  ylab("USD")                                       # <4>
```

1. Read Google's stock data from a CSV file.
2. Convert the data frame into a `tsibble.`
3. Use `autoplot()` to plot the closing prices over time.
4. Label the y-axis as "USD" for clarity.


```{r}
#| label: fig-intro-googlevol
#| fig-cap: "Daily trading volume of Google shares"
google %>%  
  as_tsibble() %>%                                   # <1>
  autoplot(Volume)                                  # <2>
```

- Convert the Google dataset into a tsibble.
- Plot the trading volume over time using `autoplot()`.

### Example: Global Temperature Data

```{r}
#| label: fig-intro-globtemp
#| message: FALSE
#| warning: FALSE
#| fig-cap: Global mean land–ocean temperature from 1850 to 2023, measured as deviations from the 1991-2020 average.

astsa::gtemp_both %>%                          # <1>
  as_tsibble() %>%                      # <2>
  autoplot() +                         # <3>
  labs(x = "Year", y = "Global Temperature Deviation (in Celsius)")  # <4>
```

1. Use `gtemp_both` dataset containing historical temperature data.
2. Convert it into a `tsibble` format.
3. Create a time series plot with `autoplot()`.
4. Label the axes to contextualize the data.


