---
title: "Activity2"
format: html
warning: false
message: false
resources:
  - template/ActivityTemplateRMarkdown.Rmd
  - template/ActivityTemplateQuarto.qmd
---

```{r}
# Time Series Essentials, install if needed!
library(feasts)       # Feature extraction & decomposition
library(fable)        # Forecasting models (ARIMA, ETS, etc.)
library(fpp3)         # Tidy time series dataseta
library(astsa)        # Applied statistical TS methods from textbook
library(tseries)      # Unit root tests & TS diagnostics
library(tsibbledata)  # Curated TS datasets
library(quantmod)     # Financial data retrieval
library(tidyquant)    # Financial analysis in tidyverse
library(purrr)        # Functional programming for TS pipelines
library(readr)        # Efficient data import
```


## Correlation, Covariance, and Partial Correlation in Time Series

In time series analysis, understanding relationships between different variables is crucial. Three fundamental statistical concepts that help capture these relationships are **covariance**, **correlation**, and **partial correlation**.

### Covariance

Covariance measures how two variables move together. For two time series $X_t$ and $Y_t$, the sample covariance over $n$ periods is given by:

$$
\text{Cov}(X, Y) = \frac{1}{n-1} \sum_{t=1}^{n} (X_t - \bar{X})(Y_t - \bar{Y})
$$

where $\bar{X}$ and $\bar{Y}$ are the sample means. A positive covariance indicates that when $X_t$ is above its mean, $Y_t$ tends to be above its mean as well, and vice versa.

### Correlation

Correlation standardizes covariance, providing a dimensionless measure between -1 and 1:

$$
\rho_{XY} = \frac{\text{Cov}(X, Y)}{\sigma_X \sigma_Y}
$$

where $\sigma_X$ and $\sigma_Y$ are the standard deviations of $X_t$ and $Y_t$. A correlation of 1 means perfect positive linear relationship, -1 means perfect negative linear relationship, and 0 suggests no linear relationship.

### Partial Correlation

Partial correlation measures the relationship between two variables while controlling for the effect of one or more additional variables. For three variables $X$, $Y$, and $Z$, the partial correlation between $X$ and $Y$ controlling for $Z$ is defined as:

$$
\rho_{XY \cdot Z} = \frac{\rho_{XY} - \rho_{XZ}\rho_{YZ}}{\sqrt{(1-\rho_{XZ}^2)(1-\rho_{YZ}^2)}}
$$

This formula removes the influence of $Z$ on both $X$ and $Y$, providing a clearer picture of their direct relationship.



## Applied Example Using `fpp3` Data

Consider the `global_economy` dataset from the `fpp3` package, which contains economic indicators for various countries over time. We can explore relationships such as:

- Covariance and correlation between Growth and CPI for different countries.
- Partial correlation to understand the relationship between Imports and Exports while controlling for GDP.

```{r}
# Filter data for Australia and select relevant variables
aus_data <- global_economy %>%
  filter(Country == "Australia") %>%    # <1>
  select(Year, GDP, Growth, Population, CPI, Imports, Exports) # <2>
```

1. Filter the `global_economy` dataset to include only rows where the country is Australia.
2. Select the variables of interest: Year, Growth, Population, and CPI



```{r}
# Calculate covariance between GDP and Population for Australia
cov_gdp_pop <- cov(aus_data$Growth, aus_data$CPI, use = "complete.obs")  # <1>
cov_gdp_pop  
```

1. Compute the covariance between Growth and CPI for the Australian subset, using complete observations.

```{r}
# Calculate correlation between GDP and Population for Australia
cor_gdp_pop <- cor(aus_data$Growth, aus_data$CPI, use = "complete.obs")  # <1>
cor_gdp_pop  
```

1. Compute the Pearson correlation coefficient between GDP and Population.

```{r}
# Compute correlation matrix for GDP, Population, and Life Expectancy
cor_matrix <- aus_data %>%
  select(GDP, Growth, Population, CPI, Imports, Exports) %>%  # <1>
  cor(use = "complete.obs")             # <2>
cor_matrix  # <11>
```

1. Select columns for GDP, Growth, Population, CPI, Imports, and Exports from the Australian data.
2. Calculate the correlation matrix for these variables using complete observations.

- Partial correlation to understand the relationship between Imports and Exports while controlling for GDP.

```{r}
# Partial correlation between Imports and Exports controlling for GDP manually
# Extract relevant correlation coefficients from the matrix
rho_Imports_Exports <- cor_matrix["Imports", "Exports"]         # <1>
rho_Imports_GDP <- cor_matrix["Imports", "GDP"]          # <2>
rho_Exports_GDP <- cor_matrix["Exports", "GDP"]  # <3>
```

1. Extracts the direct correlation between Imports and Exports.
2. Extracts the correlation between Imports and GDP.
3. Extracts the correlation between Exports and GDP.

```{r}
# Compute partial correlation between Imports and Exports controlling for GDP
pcorr_Imp_Exp_GDP <- (rho_Imports_Exports - rho_Imports_GDP * rho_Exports_GDP) /
  sqrt((1 - rho_Imports_GDP^2) * (1 - rho_Exports_GDP^2))   # <1> 
pcorr_Imp_Exp_GDP  # Display partial correlation result
```

1. Apply the partial correlation formula to compute the relationship between GDP and Life Expectancy, controlling for Population.

## Lab Activity: Recreate the above analysis using the class activity template

[Download RMarkdown Template](template/ActivityTemplateRMarkdown.Rmd){.download-link}
[Download Quarto Template](template/ActivityTemplateQuarto.qmd){.download-link}

